---
title: "Simulating Species Distributions"
format: html
editor: visual
---

This writeup is to remind me / develop a method for simulating species distributions in space and time using the methods of Thorson & Barnett 2077

The key equation is 4, which is loosely

$$
 E_p(t) \sim MVN(0, R_p \otimes V_{ep})
$$

Breaking this down, this is going to produce $E_p(t)$ the deviation from the intercept abundance, would need to make positive for marlin?

$V_{ep}$ is the correlation among species and has dimensions species x species. This one is pretty straightforward

$R_p$ is the spatial correlation between patches, and dimensions patches x patches, or from the paper "is the correlation matrix approximating similar encounter probability among locations"

This is the trickier thingy, and basically shows the rate at which locations become effectively independent in space.

This gets simulated as a Matérn function defined by the distance between each patch, a parameter $k_p$ which is the distance at which patches are essentially uncorrelated and a transformation matrix $H$ that allows decorrelation to happen faster in some direction

$$
R_p(s,s+h) = \frac{1}{2^{v-1} \Gamma(n)} \times (k_p|hH|)^n \times K_v(k_n|hH|)
$$

So, in english then. The value of the matrix R at patch p is a function of

the distance between each patch $h$

a smoothness parameter n which is fixed at 1

I have no idea what $v$ is

$k_p$ is the distance at which locations are decorrelated

$H$ is the transformation matrix with dimensions 2 X 2 and I'm not sure what the values in there should be

$K_v$ is ... something. Maybe the bessel function in R, just with a typo? But then what is $k_n$? Aha, yes per chatgpt Kv is the bessel function

`spatstat` might do it

```{r}

library(tidyverse)



resolution <- c(10,10)

distances <-
  expand_grid(x = 1:resolution[1], y = 1:resolution[2]) |>
  dist() |>
  as.matrix()

n = 1

h = seq(-100,100, by = 1)

H = 1 #matrix(rep(1,4), ncol = 2)

kp = .02

sp_cov <- rep(NA, length(h))

sigma <- 1
for (i in 1:length(h)){

sp_cov[i] = sigma^2 / (2^(n-1) * gamma(n)) * (kp * abs(h[i] * H))^n * besselK(kp*abs(h[i] * H),1)

# sp_cov[i] = (1 / (2^(n - 1) * gamma(n))) *
#     ((kp * abs(h[i]))^n) * besselK(kp * abs(h[i]), n)
# 
sp_cov[i] = rSPDE::matern.covariance(h[i],kp,1,sigma)
# 

}

# https://github.com/davidbolin/rSPDE/blob/519a51295cf1d4cfea3d98ea54ac87fba3024a9b/R/util.R#L95


plot(h, sp_cov)


rSPDE::matern.covariance(1,10,1,10)

```

https://cran.r-project.org/web/packages/spatstat/index.html

https://spatstat.org/resources.html

Let's try this out!

```{r}
#' Simulate species habitats with correlations across space and among species
#'
#' @param critters a vector of critters names of length n_species representing the number of species to be simulated
#' @param kp a rate parameter that governs the rate at which spatial cells become decorrelated with distance. Lower values mean a smoother (more correlated) habitat map 
#' @param critter_correlations leave as NA to randomly simulate correlations across species. Otherwise, a n_species x n_species correlation matrix
#' @param resolution the resolution of the system 
#' @param patch_area 
#'
#' @return
#' @export
#'
#' @examples
#' 
#' n_species <- 6
#' 
#' critters <- paste0(fruit[1:n_species],"_fish")
#' 
#' resolution <- c(4,20)
#' 
#' patch_area =  4
#' 
#' kp = .1
#' 
#' 
#' species_distributions <- sim_habitat(critters = critters, resolution = resolution, patch_area = patch_area, kp = kp)
sim_habitat <-
  function(critters,
           kp,
           critter_correlations = NA,
           resolution,
           patch_area,
           rescale_habitat = TRUE,
           max_delta = 3) {
    
    
    if (length(resolution) == 1){
      
      resolution <- rep(resolution,2)
      
    }
    
    n_species <- length(critters)
    
    patches <- prod(resolution) # the total number of patches
    
    patch_width <- sqrt(patch_area) # the width of a patch
    
    grid <-
      tidyr::expand_grid(x = 1:resolution[1], y = 1:resolution[2]) # set up spatial grid
    
    # calculate the distance
    distances <-
      grid |>
      dist() |>
      as.matrix() * patch_width
    
    # following Thorson & Barnett 2017 doi:10.1093/icesjms/fsw193
    # kp = 1e-1 # the rate parmeter; closer to 0, the more farther correlation among patches persists
    
    n = 1 # set N to 1 per paper
    
    H = 1 #matrix(rep(1,4), ncol = 2) # placeholder for now since I can't figure out how to implement anisotropy uet
    
    sigma <- 1 # placeholder which I believe should be left at 1
    
    # set up the spatial correlation matrix
    spatial_correlations <- matrix(0, nrow = (patches), ncol = (patches))
    
    for (x in 1:nrow(distances)) {
      for (y in 1:ncol(distances)) {
        if (distances[x, y] != 0) {
          spatial_correlations[x, y] <-
            sigma ^ 2 / (2 ^ (n - 1) * gamma(n)) * (kp * abs(distances[x, y] * H)) ^
            n * besselK(kp * abs(distances[x, y] * H), 1)
        } else {
          spatial_correlations[x, y] <- 1
        }
        
      }
      
    }
    
    # create species correlation matrix
    
    if (all(is.na(critter_correlations))){
      
    n_species_cores <- n_species * (n_species + 1) / 2 - n_species
    # n_species * (n_species + 1) / 2 # formula for the number of elements in the upper triangle of an n x n matric
    
    core_matrix <- matrix(0, nrow = n_species, ncol = n_species)
    
    species_cores <-
      runif(n_species_cores, min = -.1, max = .1) # randomly generate correlations among species
    # Fill in the upper triangle of the matrix
    
    core_matrix[upper.tri(core_matrix)] <- species_cores
    
    lower_triangle <- t(core_matrix)
    
    critter_correlations <- core_matrix + lower_triangle
    
    diag(critter_correlations) <- 1
    
    } 
    
    species_x_space <-
      as.matrix(Matrix::nearPD(Matrix::kronecker(spatial_correlations, critter_correlations))$mat)
    
    # generate a random species distribution for each species in space
    habitats <-
      MASS::mvrnorm(n = 1, rep(0, ncol(species_x_space)), Sigma = species_x_space)
    
    species_distributions <-
      tidyr::expand_grid(critter = critters, patch = 1:patches) |>
      cbind(grid) |>
      dplyr::arrange(patch, critter) |>
      dplyr::mutate(habitat = habitats)
    
    if (rescale_habitat){
      species_distributions$habitat <- scales::rescale(species_distributions$habitat, to = c(0, log(max_delta)))
    }
    
    check_species_cores <- species_distributions |>
      dplyr::select(patch, critter, habitat) |>
      tidyr::pivot_wider(names_from = "critter", values_from = "habitat")
    
    final_species_cores <- cor(check_species_cores[, -1])
    
    out <- list(critter_distributions = species_distributions, critter_correlations = final_species_cores )
    
  }


```

```{r}


n_species <- 10

critters <- paste0(stringr::fruit[1:n_species], "_fish")
#'
resolution <- c(4, 20)
#'
patch_area =  4
#'
kp = .08
#'
#'
species_distributions <-
  sim_habitat(
    critters = critters,
    resolution = resolution,
    patch_area = patch_area,
    kp = kp
  )

species_distributions$critter_distributions |>
  ggplot(aes(x, y, fill = (habitat))) +
  geom_tile() +
  facet_wrap( ~ critter) +
  scale_fill_viridis_c() +
  scale_x_continuous(name = "Longitude", expand = expansion(0)) +
  scale_y_continuous(name = "Latitude", expand = expansion(0))


```
