---
title: "Evaluating the Reliability of Common Indicators for Assessing Marine Protected Areas’ Contributions to Conservation and Food Security"
author:
  - name: Daniel Ovando
    affiliation:
      - name: Inter-American Tropical Tuna Commission
        department: Ecosystem & Bycatch Group
        id: 1
        address: 8901 La Jolla Shores Drive
        city: La Jolla
        state: CA
        postal-code: 92037
    email: dovando@iattc.org
    attributes:
      equal-contributor: true
      corresponding: true
  - name: Cori Lopazanski
    affiliation:
      - name: UC Santa Barbara
        department: Bren School of Environmental Science & Management
        id: 2
        address: 2400 Bren Hall
        city: Santa Barbara
        state: CA
        postal-code: 93106
    email: lopazanski@bren.ucsb.edu
    attributes:
      equal-contributor: false
      corresponding: false
format:
  docx: default
  html: 
    toc: true
    self-contained-math: true
    embed-resources: true
  pdf:
    geometry: margin=1in
    toc: true
    pdf-engine: lualatex
    include-in-header: preamble.tex
    documentclass: extarticle
    classoption: 12pt
bibliography: references.bib
abstract: |
 Something
keywords: [MPA, Simulation Modeling, Model Selection, Conservation Planning, Food Security]
execute: 
  echo: false
  warning: false
---


```{r}
library(ggplot2)
```

Hello world 

We do not report uncertainty of estimates, as our primary objective is
to evaluate relative performance across indicators rather than to make
statistical inferences. Given that this is a simulation-based analysis
with arbitrarily defined sample sizes, conventional statistical
significance testing is not meaningful [@white2014], so we do not
evaluate indicators against statistical significance thresholds (e.g., p
\< 0.05). Similarly, we do not account for statistical complexities such
as spatio-temporal clustering [@thorson2024a] or zero-inflation
[@zuur2009]. Applied use of empirical indicators would need to be
tailored to the local context, following best practices in modeling and
causal inference based on observational data from social-ecological
systems.





```{r}
#| label: fig-rho-error-bias-2
#| fig-cap: 'Spearman correlations $\rho$ (A) and $\rho^2$ (B) between indicators (y-axis) and outcomes (x-axis) at the level of individual species and fleets across all simulations. Root mean squared error (RMSE, C) and  mean error (Bias, D) between indicators (y-axis) and outcomes (x-axis) at the level of individual species and fleets across all simulations. Blue plain text indicates *Inside vs. Outside* indicators, red italic axis labels indicates a *Near vs. Far* indicator.'
#| fig-height: 7
mtcars |> 
ggplot() + 
geom_histogram(aes(mpg))


```


| **Indicator** | **Group** | **Estimated Effect & Key Assumptions** | **Example Citations** | **Equation** |   |
|------------|------------|------------|------------|------------|------------|
| Biomass Density Before-After -Control-Impact | Inside vs. Outside | Ratio of change in B after MPA implementation inside MPA relative to outside MPA. Assumes parallel trends in biomass density inside and outside MPA pre- and post-implementation | @lester2009 | $\begin{array}{l} log(B_{p,t})\sim\beta_0 + \\ - \beta_1INSIDE_p + \beta_2AFTER_t \\ +\pmb{\beta_3}INSIDE_p \times AFTER_t \end{array}$ |  |

: General structure of tested indicators of MPA effects. Bolded $\beta$
parameter in Equation column indicates parameter interpreted as the
effect of the MPA. {#tbl-test}

Testz

This is text

| **Indicator** | **Estimated Effect and Key Assumptions** | **Example Citation** | **Equation** |
|----------------|----------------|----------------|-----------------------|
| Biomass Density Before-After -Control-Impact (*Inside vs. Outside*) | Ratio of change in *B* after MPA implementation inside MPA relative to outside MPA. Assumes Parallel trends in *B* inside and outside MPA pre- and post-implementation. | @lester2009 | $\begin{array}{l} log(B_{p,t})\sim\beta_0 + \\ \beta_1INSIDE_p + \beta_2AFTER_t \\ +\pmb{\beta_3}INSIDE_p \times AFTER_t \end{array}$ |
| Biomass Density Response Ratio (*Inside vs. Outside*) | Post-MPA Ratio of *B* inside MPA relative to outside MPA. Assumes no pre-existing differences in *B* between inside and outside sites | @smith2024b | $log(B_p) \sim \beta_0 + \pmb{\beta_1}INSIDE_p$ |
| Mean Length Response Ratio (*Inside vs. Outside*) | Post-MPA ratio of *ML* inside MPA relative to outside MPA. Assumes no pre-existing differences in *ML* between inside and outside patches. | @halpern2003 | $log(ML_p) \sim \beta_0 + \pmb{\beta_1}INSIDE_p$ |
| Effort Gradient (*Near vs. Far*) | Post-MPA ratio of *E* near MPA relative to far. Assumes no pre-existing differences in *E* between near and far patches. | @goni2008 | $log(E_p) \sim \beta_0 + \pmb{\beta_1}NEAR_p$ |
| Biomass Density Gradient (*Near vs. Far*) | Post-MPA ratio of *B* near MPA relative to far from MPA. Assumes no pre-existing differences in *B* between near and far patches. | @halpern2009 | $log(B_p) \sim \beta_0 + \pmb{\beta_1}NEAR_p$ |
| Biomass Density Before-After Gradient (*Near vs. Far*) | Ratio of change in *B* after MPA implementation near to MPA relative to far from MPA. Assumes Parallel trends in *B* density near and far from MPA pre- and post-implementation. | @lynham2024 | $\begin{array}{l} log(B_{p,t})\sim \beta_0 + \\ \beta_1NEAR_p + \beta_2AFTER_t + \\ \pmb{\beta_3}NEAR_p \times AFTER_t \end{array}$ |

: General structure of tested indicators of MPA effects. Bolded $\beta$
parameter in **Equation** column indicates parameter interpreted as the
effect of the MPA. *B* refers to biomass, *E* refers to effort, *ML*
refers to mean length. *INSIDE* refers to patches *p* located inside MPA
borders (whether before or after MPA implementation), *OUTSIDE* to
patches outside. *AFTER* refers to time periods *t* after MPA
implementation, *BEFORE* is before MPA implementation. *NEAR* are
patches *p* outside but near the border of an MPA (whether before or
after MPA implementation), FAR is outside and farther from the border of
an MPA. {#tbl-indicators}
